preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-14 Manuscript/manuscript.Rmd

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

@@

$$N_{e} = \dfrac{1}{b} \Sigma_{i=1}^{n}e_{i}$$
where $e_{i}$ is the number ChIP-seq peak regions that overlap with the $i^{th}$ predicted boundary, $b$, flanked by resolution $r$. We evaluated $N_{e}$ for combinations of $t=\{0.95, 0.975, 0.99, 1.0 \}$ and $\epsilon = \{5000, 10000, 15000, 20000, 25000 \}$. The heuristic of $\epsilon$ is that density-reachable bases with genomic distances less than $\epsilon$ should occupy the same designated cluster.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [???refer to supp tables 4 and 5 here]

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across 10 kb, 25 kb, 50 kb, and 100 kb data resolutions for both the (A) GM12878 and (B) K562 cell lines, respectively. Distances are measured as the number of bases from the center of a genomic bin to the center of the nearest genomic annotation. Each density curve represents individual genomic annotation.

>>>>

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

@@

$$N_{e} = \dfrac{1}{b} \Sigma_{i=1}^{n}e_{i}$$

where $e_{i}$ is the number ChIP-seq peak regions that overlap with the $i^{th}$ predicted boundary, $b$, flanked by resolution $r$. We evaluated $N_{e}$ for combinations of $t=\{0.95, 0.975, 0.99, 1.0 \}$ and $\epsilon = \{5000, 10000, 15000, 20000, 25000 \}$. The heuristic of $\epsilon$ is that density-reachable bases with genomic distances less than $\epsilon$ should occupy the same designated cluster.

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us to investigate its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [???refer to supp tables 4 and 5 here when ready]

@@

**Supplementary Figure 1.** Density curves illustrate the normalization of distances before (red) and after (blue) performing a $log_{2}$ transformation across 10 kb, 25 kb, 50 kb, and 100 kb data resolutions for both the (A) GM12878 and (B) K562 cell lines, respectively. Distances are measured as the number of bases from the center of a genomic bin to the center of the nearest genomic annotation. All genomic annotations were used. Each density curve represents an individual genomic annotation.
